---
title: "Does Urban Vegetation Mitigate High Temperatures Across Cities?"
author: "Shiyi Zheng & Juanying Huangfu"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load packages
library(tidyverse)
library(lubridate)
library(sf)
library(here)
library(ggplot2)

```

# 1. Introduction

China’s rapid urbanization over the past decades has significantly intensified the Urban Heat Island effect (UHI), leading to rising urban temperatures and increasing concerns over heat-related health risks, energy demand, and long-term urban sustainability. Empirical studies have shown that urban areas in China have experienced faster warming trends compared to surrounding rural regions, largely driven by land-use change and the expansion of impervious surfaces (Zhou et al., 2014).Urban vegetation, commonly measured using the Normalized Difference Vegetation Index (NDVI), has been widely recognized as an effective nature-based solution to mitigate urban heat. Vegetation can reduce surface and air temperatures through mechanisms such as shading and evapotranspiration, thereby improving thermal comfort and reducing heat exposure(Bowler et al., 2010). And remote sensing studies using NDVI have consistently found negative associations between vegetation cover and land surface temperature across urban environments(Peng et al., 2011).

However, despite this well-established relationship, the extent to which vegetation mitigates temperature across Chinese cities over time remains insufficiently quantified. Most existing studies focus on single cities or short time periods, limiting the ability to capture long-term dynamics and cross-city variability. Moreover, regional heterogeneity—such as differences in climate zones, urban form, and baseline vegetation coverage—may lead to substantial variation in the strength and direction of the NDVI–temperature relationship(Zhao et al., 2014).

To address these gaps, this study conducts a multi-city analysis across Chinese prefecture-level cities from 2000 to 2024, integrating NDVI and temperature data to examine both temporal trends and cross-sectional relationships. Specifically, the analysis explores (1) long-term changes in vegetation and temperature, (2) the overall association between NDVI and temperature, (3) regional differences in this relationship, and (4) whether changes in vegetation are associated with corresponding changes in temperature over time. By combining time series analysis, regression approaches, and spatial visualization, this study aims to provide a systematic comprehensive understanding of the role of urban vegetation in mitigating heat across rapidly urbanizing regions.

**Research Question:**

-   How does vegetation cover (NDVI) affect temperature across Chinese cities over time?

**Sub-questions:**

-   How have NDVI and temperature changed over time?
-   Is higher vegetation associated with lower temperatures?
-   Does the relationship vary across regions?
-   How does temperature differ across vegetation levels?
-   Are changes in vegetation associated with changes in temperature?

------------------------------------------------------------------------

# 2. Hypotheses

# 3. Data

## 3.1 Data Sources

This study uses two main datasets.

Vegetation data are measured using annual NDVI for Chinese prefecture-level cities from 2000 to 2024. These data are derived from MODIS satellite products and aggregated to the city level.

Temperature data consist of annual average temperature at the city level over the same period. These data are obtained from gridded climate datasets and spatially aggregated to match administrative boundaries.

## 3.2 Data Processing

The NDVI data are originally based on monthly raster observations and are aggregated into annual averages. Both NDVI and temperature datasets are spatially averaged within prefecture-level city boundaries using administrative shapefiles.

The two datasets are then merged by city and year to construct a consistent panel dataset.

## 3.3 Final Dataset

The final dataset is a city–year panel covering the period 2000 to 2024.

Key variables include:

-   NDVI (annual average vegetation index)
-   Temperature (annual average, °C)
-   City identifier
-   Year
-   Province

------------------------------------------------------------------------

# 4. Data Wrangling

In this step, the NDVI and temperature datasets are cleaned, standardized, and merged into a unified panel dataset.

The main steps include:

-   Standardizing variable formats such as city names and year
-   Converting variables into appropriate data types
-   Merging NDVI and temperature data by city and year
-   Removing missing or inconsistent observations
-   Ensuring a consistent time period across cities

The resulting dataset forms the basis for all subsequent analysis.

```{r}
# Data import and cleaning steps

# check project root
here()

# create processed folder if it does not exist
dir.create(here("data", "processed"), showWarnings = FALSE, recursive = TRUE)

# 1. Read raw data

temp_raw <- read.csv(
  here("data", "raw", "clty level_temp_2000-2024.csv")
)

ndvi_raw <- read.csv(
  here("data", "raw", "city level_ndvi_2000-2024.csv")
)

# 2. Define year columns

year_cols <- as.character(2000:2024)

# 3. Reshape temperature data to long format

temp_long <- temp_raw %>%
  clean_names_basic() %>%
  pivot_longer(
    cols = any_of(year_cols),
    names_to = "year",
    values_to = "temperature"
  ) %>%
  mutate(
    year = as.integer(year),
    temperature = as.numeric(temperature)
  ) %>%
  arrange(city_id, year)

# 4. Reshape NDVI data to long format

ndvi_long <- ndvi_raw %>%
  clean_names_basic() %>%
  pivot_longer(
    cols = any_of(year_cols),
    names_to = "year",
    values_to = "ndvi"
  ) %>%
  mutate(
    year = as.integer(year),
    ndvi = as.numeric(ndvi)
  ) %>%
  arrange(city_id, year)


# 5. Check duplicate city-year observations

panel_check_temp <- temp_long %>%
  count(city_id, year) %>%
  filter(n > 1)

panel_check_ndvi <- ndvi_long %>%
  count(city_id, year) %>%
  filter(n > 1)

# 6. Merge temperature and NDVI

processed_city_panel <- temp_long %>%
  left_join(
    ndvi_long,
    by = c("city_id", "city", "pr_id", "province", "year")
  ) %>%
  mutate(
    temp_anomaly = temperature - mean(temperature, na.rm = TRUE),
    ndvi_z = as.numeric(scale(ndvi)),
    temperature_z = as.numeric(scale(temperature))
  ) %>%
  select(
    city_id, city, pr_id, province, year,
    ndvi, temperature, ndvi_z, temperature_z, temp_anomaly
  ) %>%
  arrange(city_id, year)


# 7. Summary before dropping missing values

missing_summary_before <- processed_city_panel %>%
  summarise(
    n_rows = n(),
    n_cities = n_distinct(city_id),
    year_min = min(year, na.rm = TRUE),
    year_max = max(year, na.rm = TRUE),
    missing_ndvi = sum(is.na(ndvi)),
    missing_temperature = sum(is.na(temperature))
  )

city_year_completeness_before <- processed_city_panel %>%
  group_by(city_id, city) %>%
  summarise(
    n_years = n(),
    missing_ndvi = sum(is.na(ndvi)),
    missing_temperature = sum(is.na(temperature)),
    .groups = "drop"
  )


# 8. Drop rows with missing values


processed_city_panel_clean <- processed_city_panel %>%
  drop_na(ndvi, temperature)


# 9. Summary after dropping missing values


missing_summary_after <- processed_city_panel_clean %>%
  summarise(
    n_rows = n(),
    n_cities = n_distinct(city_id),
    year_min = min(year, na.rm = TRUE),
    year_max = max(year, na.rm = TRUE),
    missing_ndvi = sum(is.na(ndvi)),
    missing_temperature = sum(is.na(temperature))
  )

city_year_completeness_after <- processed_city_panel_clean %>%
  group_by(city_id, city) %>%
  summarise(
    n_years = n(),
    .groups = "drop"
  )


# 10. Export processed files

write_csv(
  processed_city_panel_clean,
  here("data", "processed", "processed_city_ndvi_temp_clean.csv")
)

write_csv(
  missing_summary_before,
  here("data", "processed", "processed_city_ndvi_temp_summary_before_dropna.csv")
)

write_csv(
  missing_summary_after,
  here("data", "processed", "processed_city_ndvi_temp_summary_after_dropna.csv")
)

write_csv(
  city_year_completeness_before,
  here("data", "processed", "processed_city_ndvi_temp_completeness_before_dropna.csv")
)

write_csv(
  city_year_completeness_after,
  here("data", "processed", "processed_city_ndvi_temp_completeness_after_dropna.csv")
)

saveRDS(
  processed_city_panel_clean,
  here("data", "processed", "processed_city_ndvi_temp_clean.rds")
)

# 11. Print checks

print(missing_summary_before)
print(missing_summary_after)
print(panel_check_temp)
print(panel_check_ndvi)
```

# 5. Exploratory Data Analysis

## 5.1 Temporal Trends

As the first step in the exploratory data analysis, this study examines how NDVI and temperature evolve over time by aggregating city-level observations into annual averages. This transformation simplifies the panel dataset into a time series representation, allowing for a clear overview of long-term dynamics.

Line plots are used to visualize temporal changes in both variables, as they effectively capture year-to-year variation while preserving overall trends. To support interpretation, locally estimated scatterplot smoothing (LOESS) curves are applied to highlight underlying patterns by reducing short-term fluctuations.

The purpose of this step is not to establish relationships between variables, but to provide a descriptive overview of their temporal behavior. By identifying whether NDVI and temperature exhibit increasing, decreasing, or stable trends over the study period, this analysis lays the foundation for subsequent exploration of distributions (Section 5.2) and cross-city heterogeneity (Section 5.3).

```{r}
# 1. Read Data
library(readr)
processed_city_ndvi_temp_clean <- read_csv("data/processed/processed_city_ndvi_temp_clean.csv")

# 2. Check Data
View(processed_city_ndvi_temp_clean)
glimpse(processed_city_ndvi_temp_clean)
summary(processed_city_ndvi_temp_clean[, c("year", "ndvi", "temperature")])

#Check whether there are sufficient urban observations every year
processed_city_ndvi_temp_clean %>%
  group_by(year) %>%
  summarise(
    n_city = n_distinct(city),
    n_obs = n()
  )

```

The dataset contains city-level observations of NDVI and temperature from 2000 to 2024. Inspection of the data structure confirms that both NDVI and temperature are recorded as continuous numeric variables, alongside identifiers such as city and year.

Summary statistics show that NDVI values range from 0.060 to 0.829, with a mean of 0.478 and a median of 0.495. Temperature values range from -5.604°C to 27.053°C, with a mean of 13.486°C and a median of 15.140°C.

The yearly observation counts indicate a consistent number of cities (369) and total observations (374) across all years. This suggests that the dataset is balanced over time, with no missing years or substantial variation in the number of observations.

```{r}
# 3. Aggregate by year
ts_data <- processed_city_ndvi_temp_clean %>%
  group_by(year) %>%
  summarise(
    mean_ndvi = mean(ndvi, na.rm = TRUE),
    mean_temp = mean(temperature, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
# 4. Temporal Trends of NDVI
ggplot(ts_data, aes(x = year, y = mean_ndvi)) +
  geom_line(color = "darkgreen", size = 1) +
  geom_smooth(method = "loess", color = "darkgrey", se = FALSE) +
  labs(
    title = "Temporal Trend of NDVI Across Cities in China (2000–2024)",
    x = "Year",
    y = "Mean NDVI",
  ) +
  theme_minimal()
```

The temporal trend of NDVI shows a clear and steady increase over the study period from 2000 to 2024. The mean NDVI rises from approximately 0.44 in 2000 to above 0.50 by 2024, indicating a gradual upward trend.

While minor year-to-year fluctuations are observed—particularly in the early 2000s and around 2008–2010—the overall trajectory remains consistently increasing. The LOESS smoothed curve further confirms this pattern, showing a continuous upward trend without major reversals.

Overall, the results suggest that average vegetation levels across cities have increased over time, with relatively stable growth throughout the study period.

```{r}
# 5. Temporal Trends of Temperature
ggplot(ts_data, aes(x = year, y = mean_temp)) +
  geom_line(color = "darkred", size = 1) +
  geom_smooth(method = "loess", color = "darkgrey", se = FALSE) +
  labs(
    title = "Temporal Trend of Temperature Across Cities in China (2000–2024)",
    x = "Year",
    y = "Mean Temperature (°C)",
  ) +
  theme_minimal()
```

The temporal trend of temperature shows an overall increasing pattern from 2000 to 2024. The mean temperature rises from approximately 13.0°C in 2000 to above 14.0°C by the end of the study period.

Compared to NDVI, temperature exhibits more pronounced year-to-year fluctuations, particularly between 2000 and 2012, where several short-term increases and decreases are observed. Despite this variability, the overall trend remains upward.

The LOESS smoothed curve highlights a relatively stable or slightly declining period in the early years, followed by a clearer and more consistent increase after around 2012. This indicates that the warming trend becomes more pronounced in the later period.

## 5.2 Distribution of Key Variables

Following the examination of temporal trends, the analysis next focuses on the distribution of key variables across all observations. Understanding the distributional properties of NDVI and temperature is essential for interpreting the overall data structure before conducting further analysis.

Histograms are used as a primary visualization tool to assess how values of NDVI and temperature are distributed across the dataset. These plots provide insights into the range, central tendency, and dispersion of each variable, while also revealing potential skewness or the presence of extreme values. In addition, summary statistics are used to complement the visual analysis by quantifying key characteristics such as mean, median, and variability.

This step serves as a diagnostic check on the dataset. It helps ensure that the variables behave as expected, identifies any potential outliers, and informs subsequent analyses, particularly regression and comparative analysis, where distributional assumptions and data balance may influence results.

```{r}
# 1. Reshape for combined exploratory plots
dist_data <- processed_city_ndvi_temp_clean %>%
  select(ndvi, temperature) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  )
```

```{r}
# 2. Histograms with density curves
ggplot(dist_data, aes(x = value, fill = variable)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 color = "white",
                 alpha = 0.6) +
  geom_density(color = "black", linewidth = 1) +
  facet_wrap(~ variable, scales = "free") +
  labs(
    title = "Distributions of NDVI and Temperature",
    x = "Value",
    y = "Density"
  ) +
  theme_minimal()
```

The distributions of NDVI and temperature exhibit distinct patterns across all observations.

NDVI values are relatively concentrated within a narrower range, with most observations clustered between approximately 0.4 and 0.7. The density curve shows a slightly multimodal shape, indicating the presence of multiple concentration points within the mid-range values.

In contrast, temperature values display a wider spread, ranging from below 0°C to above 20°C. The distribution is more dispersed, with a dominant peak around the mid-teens, and a longer tail extending toward lower temperatures. The density curve also suggests multiple local peaks, reflecting variation across observations.

Overall, the comparison indicates that NDVI is more tightly distributed, while temperature exhibits greater variability and dispersion.

```{r}
# 3. Boxplots for dispersion and potential outliers
ggplot(dist_data, aes(x = variable, y = value, fill = variable)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red") +
  facet_wrap(~ variable, scales = "free_y")+
  labs(
    title = "Boxplots of NDVI and Temperature",
    x = "Variable",
    y = "Value"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The boxplots provide a clearer view of the dispersion and range of NDVI and temperature values.

NDVI values are relatively concentrated, with most observations falling within a narrower interquartile range centered around approximately 0.4 to 0.6. The overall range extends to both lower and higher values, indicating some variation across observations, but the distribution remains relatively compact.

In contrast, temperature exhibits a much wider spread. The interquartile range is larger, and the whiskers extend over a broader range compared to NDVI. A few observations appear below the lower whisker, indicating the presence of lower-end values outside the main distribution.

Overall, the boxplots show that temperature has greater dispersion than NDVI, while NDVI values remain more tightly clustered.

```{r}
# 4. Separate histogram for NDVI
ggplot(processed_city_ndvi_temp_clean, aes(x = ndvi)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 fill = "darkgreen",
                 color = "white",
                 alpha = 0.8) +
  geom_density(color = "black", linewidth = 1) +
  labs(
    title = "Distribution of NDVI Across All Observations",
    x = "NDVI",
    y = "Density"
  ) +
  theme_minimal()
```

The distribution of NDVI values is relatively concentrated within the mid-range, with most observations falling between approximately 0.4 and 0.7. The density curve indicates a slightly multimodal pattern, with two noticeable peaks within this interval.

Lower NDVI values (below 0.3) are less frequent, and only a small number of observations extend toward higher values above 0.7. Overall, the distribution suggests that most cities exhibit moderate levels of vegetation coverage, with limited extreme values.

```{r}
# 5. Separate histogram for Temperature
ggplot(processed_city_ndvi_temp_clean, aes(x = temperature)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 fill = "darkred",
                 color = "white",
                 alpha = 0.8) +
  geom_density(color = "black", linewidth = 1) +
  labs(
    title = "Distribution of Temperature Across All Observations",
    x = "Temperature (°C)",
    y = "Density"
  ) +
  theme_minimal()

```

The distribution of temperature values is more widely spread compared to NDVI, covering a broad range from below 0°C to above 20°C. Most observations are concentrated between approximately 8°C and 18°C, with a prominent peak around the mid-teens.

The density curve indicates a multimodal pattern, with additional smaller peaks at lower and higher temperature ranges. The left tail extends toward lower temperature values, suggesting the presence of relatively low-temperature observations, while a smaller right tail reaches above 20°C.

Overall, the distribution shows greater variability and dispersion than NDVI, with a wider range and more pronounced variation across observations.

## 5.3 City-Level Variation

While the previous sections examine overall temporal trends and the distribution of key variables, these aggregated perspectives may mask important differences across cities. To better understand the extent of heterogeneity, this section explores city-level variation in NDVI and temperature over time. By visualizing individual city trajectories, the analysis aims to assess whether observed aggregate patterns are consistent across cities or driven by a subset of observations.

```{r}
# 1. Check number of cities and observations per city
processed_city_ndvi_temp_clean %>%
  summarise(
    total_cities = n_distinct(city),
    total_obs = n()
  )

processed_city_ndvi_temp_clean %>%
  group_by(city) %>%
  summarise(n_years = n()) %>%
  summary()
```

The dataset includes observations from 369 distinct cities. At the city level, most cities have 25 years of data, as indicated by the minimum, first quartile, and median values of 25 years. This suggests that the majority of cities are consistently observed throughout the study period from 2000 to 2024.

The mean number of observations per city is slightly higher (25.34), and the maximum reaches 50, indicating that a small number of cities have more frequent observations than others. Overall, the dataset is largely balanced across cities, with only limited variation in the number of observations per city.

```{r}
# 2. Overall trajectory across cities 

# NDVI trajectories across cities 
ggplot(processed_city_ndvi_temp_clean,
       aes(x = year, y = ndvi, group = city)) +
  geom_line(alpha = 0.1, color = "darkgreen") +
  geom_smooth(aes(group = 1), method = "loess",
              color = "black", linewidth = 1.2) +
  labs(
    title = "City-Level NDVI Trajectories (2000–2024)",
    x = "Year",
    y = "NDVI"
  ) +
  theme_minimal()

# Temperature trajectories across cities
ggplot(processed_city_ndvi_temp_clean,
       aes(x = year, y = temperature, group = city)) +
  geom_line(alpha = 0.1, color = "darkred") +
  geom_smooth(aes(group = 1), method = "loess",
              color = "black", linewidth = 1.2) +
  labs(
    title = "City-Level Temperature Trajectories (2000–2024)",
    x = "Year",
    y = "Temperature (°C)"
  ) +
  theme_minimal()
```

The city-level trajectories reveal substantial heterogeneity across cities for both NDVI and temperature. Individual city trends, shown by the semi-transparent lines, vary widely in both level and variation over time.

The city-level NDVI trajectories show clear variation across cities. From the figure, most NDVI values appear to fall roughly between 0.2 and 0.8, while the smoothed trend line gradually increases from about 0.44 in 2000 to above 0.50 by 2024.Although the aggregate smoothed trend is upward, the individual city lines are spread across different NDVI levels. This suggests that average vegetation conditions differ across cities and that the overall trend does not fully represent every city’s trajectory.

The city-level temperature trajectories show wider vertical dispersion than NDVI. From the figure, temperature values range broadly from below 0°C to above 20°C for many city-year observations. The smoothed trend line increases gradually over time, especially after the early 2010s.The spread of individual lines indicates that cities differ substantially in their temperature levels. Therefore, the average warming pattern should be interpreted as an overall tendency rather than a uniform trend across all cities.

Overall, these results indicate that aggregate trends may mask substantial differences across cities. The observed heterogeneity highlights the importance of examining city-level variation rather than relying solely on averaged patterns.

```{r}
# 3. Sampling city avoid over plotting
set.seed(123)

sample_cities <- processed_city_ndvi_temp_clean %>%
  distinct(city) %>%
  sample_n(12)

sample_data <- processed_city_ndvi_temp_clean %>%
  filter(city %in% sample_cities$city)

# Sampled NDVI
ggplot(sample_data, aes(x = year, y = ndvi, color = city)) +
  geom_line() +
  labs(
    title = "NDVI Trends for Sampled Cities",
    x = "Year",
    y = "NDVI"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Sampled Temperature
ggplot(sample_data, aes(x = year, y = temperature, color = city)) +
  geom_line() +
  labs(
    title = "Temperature Trends for Sampled Cities",
    x = "Year",
    y = "Temperature (°C)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

To further examine city-level variation, a subset of cities is selected for detailed visualization. Compared to the aggregated view, these plots make individual trajectories more clearly observable.

For NDVI, all sampled cities show a generally increasing trend over time, but the levels and rates of increase differ. Some cities start at higher NDVI levels and maintain relatively high values throughout the period, while others remain at lower levels despite gradual increases. This indicates that although the overall upward trend is consistent, the magnitude and baseline of vegetation coverage vary across cities.

For temperature, a similar pattern of heterogeneity is observed. All sampled cities exhibit relatively stable or gradually increasing trends, but there are clear differences in baseline temperature levels. Some cities consistently experience higher temperatures, while others remain at lower levels, with relatively parallel trajectories over time.

Overall, these plots reinforce that while general trends exist, substantial differences across cities persist in both NDVI and temperature.

# 6. Methods and Analysis

## 6.1 Time Series Analysis

This section examines long-term trends in NDVI and temperature over time.
Following the logic of the course time series exercises, we first aggregate the panel data to yearly averages, then use line plots and smoothed trend lines to visualize temporal patterns.
Because our dataset is annual rather than monthly or daily, the main focus here is long-term trend detection rather than seasonal decomposition.

```{r}
data_clean <- read.csv(
  here("data", "processed", "processed_city_ndvi_temp_clean.csv"),
  stringsAsFactors = FALSE
)

# Aggregate city-year panel data to annual mean values across all cities
annual_summary <- data_clean %>%
  group_by(year) %>%
  summarise(
    mean_ndvi = mean(ndvi, na.rm = TRUE),
    mean_temperature = mean(temperature, na.rm = TRUE),
    sd_ndvi = sd(ndvi, na.rm = TRUE),
    sd_temperature = sd(temperature, na.rm = TRUE),
    n_cities = n_distinct(city),
    .groups = "drop"
  )

annual_summary
```

The aggregated dataset summarizes the average vegetation and temperature conditions across Chinese prefecture-level cities for each year from 2000 to 2024.

```{r}
# 1. Smoothed trend plots
# NDVI Trend Over Time
ggplot(annual_summary, aes(x = year, y = mean_ndvi)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 1) +
  labs(
    title = "Average NDVI Across Chinese Cities, 2000–2024",
    x = "Year",
    y = "Mean NDVI"
  )

# Temperature Trend Over Time
ggplot(annual_summary, aes(x = year, y = mean_temperature)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 1) +
  labs(
    title = "Average Temperature Across Chinese Cities, 2000–2024",
    x = "Year",
    y = "Mean Temperature (°C)"
  )
```

The time series plots show a clear upward trend in both NDVI and temperature from 2000 to 2024, but the patterns are not identical.

NDVI increases quite steadily over time.
There are small fluctuations in the early 2000s, but after around 2010 the increase becomes more consistent, and the overall trend is very smooth.
This suggests that vegetation coverage in Chinese cities has been gradually improving over the past two decades.

Temperature also shows an overall increasing trend, but the pattern is more variable.
There is a noticeable dip around 2010–2012, followed by a steady increase afterward.
Compared to NDVI, temperature changes are less smooth and show more short-term fluctuations, which is expected given the influence of climate variability.

```{r}
# 2. Time series object representation
# Create annual time series objects
ndvi_ts <- ts(
  annual_summary$mean_ndvi,
  start = min(annual_summary$year),
  frequency = 1
)

temp_ts <- ts(
  annual_summary$mean_temperature,
  start = min(annual_summary$year),
  frequency = 1
)

# Plot time series objects
plot(
  ndvi_ts,
  main = "NDVI Time Series (Annual Average)",
  xlab = "Year",
  ylab = "Mean NDVI"
)

plot(
  temp_ts,
  main = "Temperature Time Series (Annual Average)",
  xlab = "Year",
  ylab = "Mean Temperature (°C)"
)
```

These time series plots show the same patterns in a more direct way, without smoothing.
The upward trends in both NDVI and temperature are clearly visible, although year-to-year variation is more noticeable compared to the smoothed plots.

```{r}
# 3. Mann-Kendall Trend Tests
ndvi_mk <- mk.test(ndvi_ts)
temp_mk <- mk.test(temp_ts)

ndvi_mk
temp_mk
```

The Mann-Kendall test results support what we observe in the plots.
NDVI shows a very strong and statistically significant increasing trend (z = 6.19, p \< 0.001, τ = 0.89), indicating that the upward trend is highly consistent over time.

Temperature also shows a statistically significant increasing trend (z = 3.53, p \< 0.001, τ = 0.51), although the strength of the trend is more moderate compared to NDVI.

One important takeaway is that both vegetation and temperature are increasing at the same time.
This is directly relevant to our research question.

If vegetation has a cooling effect, we might expect higher NDVI to be associated with lower temperatures.
However, the fact that both variables increase over time suggests that there may be broader factors, such as climate change, driving temperature upward.

This also highlights an important issue for the next step of the analysis: we need to separate the effect of vegetation from overall time trends, rather than relying only on simple correlations.

## 6.2 Correlation and Linear Regression: Relationship Between NDVI and Temperature

We next examine whether higher vegetation coverage is associated with lower temperature.
Following the logic of the course GLM exercises, we first visualize the relationship using a scatter plot and a fitted linear trend line.
We then use correlation analysis and a simple linear regression model to quantify the direction and strength of the association between NDVI and temperature.

```{r}
ndvi_temp <- data_clean %>%
  select(city, year, ndvi, temperature) %>%
  drop_na()

glimpse(ndvi_temp)
summary(ndvi_temp)
```

```{r}
# 1. Scatter plot with fitted linear trend
ggplot(ndvi_temp, aes(x = ndvi, y = temperature)) +
  geom_point(alpha = 0.3, size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = "Temperature and NDVI Across Chinese Cities",
    x = "NDVI",
    y = "Temperature (°C)"
  )
```

The scatter plot shows a clear positive relationship between NDVI and temperature.
As NDVI increases, temperature also tends to increase, and the fitted regression line has a strong upward slope.

This pattern is quite striking.
Instead of seeing a cooling effect of vegetation, the plot suggests that cities with higher NDVI tend to have higher temperatures on average.
The points are fairly spread out, but the overall upward trend is very consistent across the full range of NDVI values.

```{r}
# 2.Pearson correlation test
cor_test_ndvi_temp <- cor.test(ndvi_temp$ndvi, ndvi_temp$temperature)
cor_test_ndvi_temp
```

The Pearson correlation coefficient is about 0.69, which indicates a relatively strong positive relationship between NDVI and temperature.
The p-value is extremely small (p \< 0.001), meaning this relationship is statistically significant and very unlikely to be due to random chance.

In practical terms, this means that NDVI and temperature tend to move together across observations.
Cities and years with higher vegetation levels also tend to have higher temperatures.

```{r}
# 3.Simple linear regression
ndvi_temp_lm <- lm(temperature ~ ndvi, data = ndvi_temp)
summary(ndvi_temp_lm)
```

The regression results are consistent with the scatter plot and correlation test.
The positive slope indicates that, on average, higher NDVI is associated with higher temperature.

This result is quite different from what we might expect if vegetation had a simple cooling effect.
Instead of a negative relationship, we observe a strong positive association.
This suggests that the relationship between vegetation and temperature is more complex than a simple “more trees, lower temperature” story.

## 6.3 Regional Comparison

In this section, we explore whether the relationship between NDVI and temperature differs across regions in China.
To do this, we classify cities into major geographic regions (e.g., North China, East China, South China) based on their province and then compare patterns across these groups.

```{r}
# 1.  create region classification based on province
data_clean <- data_clean %>%
  mutate(
    region = case_when(
      # North China
      province %in% c(
        "Beijing Municipality",
        "Tianjin Municipality",
        "Hebei Province",
        "Shanxi Province",
        "Inner Mongolia Autonomous Region"
      ) ~ "North China",

      # East China
      province %in% c(
        "Shanghai Municipality",
        "Jiangsu Province",
        "Zhejiang Province",
        "Anhui Province",
        "Fujian Province",
        "Jiangxi Province",
        "Shandong Province",
        "Taiwan Province"
      ) ~ "East China",

      # Central China
      province %in% c(
        "Henan Province",
        "Hubei Province",
        "Hunan Province"
      ) ~ "Central China",

      # South China
      province %in% c(
        "Guangdong Province",
        "Guangxi Zhuang Autonomous Region",
        "Hainan Province",
        "Hong Kong SAR",
        "Macau SAR"
      ) ~ "South China",

      # Southwest
      province %in% c(
        "Chongqing Municipality",
        "Sichuan Province",
        "Guizhou Province",
        "Yunnan Province",
        "Tibet Autonomous Region"
      ) ~ "Southwest",

      # Northwest
      province %in% c(
        "Shaanxi Province",
        "Gansu Province",
        "Qinghai Province",
        "Ningxia Hui Autonomous Region",
        "Xinjiang Uygur Autonomous Region"
      ) ~ "Northwest",

      # Northeast
      province %in% c(
        "Liaoning Province",
        "Jilin Province",
        "Heilongjiang Province"
      ) ~ "Northeast",

      TRUE ~ "Other"
    )
  )
table(data_clean$region)
```

```{r}
# 2. scatter plot by region
ggplot(data_clean, aes(x = ndvi, y = temperature)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  facet_wrap(~ region) +
  labs(
    title = "NDVI and Temperature by Region",
    x = "NDVI",
    y = "Temperature (°C)"
  )
```

The relationship between NDVI and temperature clearly differs across regions, and the patterns are not uniform.

In East China, North China, and Southwest China, the relationship is strongly positive.
In these regions, higher NDVI is consistently associated with higher temperature, and the slope is quite steep.
This is especially clear in Southwest China, where the trend line increases sharply as NDVI increases.

In contrast, the Northeast shows a different pattern.
The relationship there is slightly negative, suggesting that higher NDVI may be associated with lower temperature in that region.
This is the only region where the pattern looks closer to the expected “cooling effect” of vegetation.

Other regions, such as Northwest and Central China, show weaker or more mixed patterns.
The points are more scattered, and the slope is flatter, suggesting that the relationship is less stable.

Overall, the figure suggests that the NDVI–temperature relationship is highly region-dependent, rather than following a single national pattern.

```{r}
# 3. interaction regression
region_lm <- lm(temperature ~ ndvi * region, data = data_clean)
summary(region_lm)
```

The regression results confirm what we see in the plots.
The interaction terms between NDVI and region are mostly significant, which means the slope of NDVI is different across regions.

For the baseline region (Central China), the coefficient on NDVI is small and not statistically significant, suggesting that NDVI is not strongly related to temperature there.

However, once we look at other regions, the pattern changes.
In East China and North China, the interaction terms are large and positive, indicating that NDVI has a much stronger positive association with temperature in these regions.
In Southwest China, the effect is extremely strong, with a very steep increase in temperature as NDVI rises.

In contrast, the Northeast shows a negative interaction effect, meaning that the relationship between NDVI and temperature becomes negative in that region.
This aligns with the visual pattern in the scatter plot.

Overall, the regression results suggest that the relationship between vegetation and temperature is not only non-uniform, but can even change direction depending on the region.

One important takeaway is that regional climate and geography seem to play a major role in shaping this relationship.

In warmer regions, such as South China and Southwest China, higher NDVI tends to coincide with higher temperatures.
This likely reflects underlying climate conditions, where both vegetation and temperature are naturally higher.

In colder regions, such as the Northeast, the relationship appears closer to what we might expect, with higher vegetation potentially linked to lower temperature.

This suggests that the simple relationship between NDVI and temperature cannot be interpreted in isolation.
Instead, it is strongly influenced by broader environmental conditions.
This also reinforces the idea from the previous section that we need to be careful when interpreting the NDVI–temperature relationship, as it may be driven by shared geographic and climatic factors rather than a direct causal effect.

## 6.4 Distribution Analysis

We further examine how temperature varies across different levels of vegetation coverage.
NDVI values are grouped into quartiles, and temperature distributions are compared across these groups using boxplots.
This approach allows us to assess whether cities with higher vegetation consistently experience lower temperatures.

```{r}
# 1. grouping by NDVI
ndvi_grouped <- data_clean %>%
  mutate(
    ndvi_group = ntile(ndvi, 4)  # 分成4组（quartiles）
  )

ndvi_grouped <- ndvi_grouped %>%
  mutate(
    ndvi_group = factor(
      ndvi_group,
      levels = c(1, 2, 3, 4),
      labels = c("Low NDVI", "Lower-Mid NDVI", "Upper-Mid NDVI", "High NDVI")
    )
  )
```

```{r}
# 2.boxplot
ggplot(ndvi_grouped, aes(x = ndvi_group, y = temperature)) +
  geom_boxplot(outlier.alpha = 0.2) +
  geom_jitter(alpha = 0.1, width = 0.2) +
  labs(
    title = "Temperature Distribution Across NDVI Levels",
    x = "NDVI Group",
    y = "Temperature (°C)"
  )
```

The boxplots show a clear upward shift in temperature as NDVI increases.

From the low NDVI group to the high NDVI group, both the median temperature and the overall distribution move higher.
In particular, the median temperature increases steadily across the four groups, and the upper range of temperature is also noticeably higher in the high NDVI group.

This pattern is quite consistent across the data.
Even though there is still variation within each group, the overall trend is very clear: cities with higher NDVI tend to have higher temperatures rather than lower ones.

This result reinforces what we observed in the previous analyses.
Instead of finding evidence that higher vegetation leads to cooler temperatures, we again see a strong positive association between NDVI and temperature.

One important point is that this pattern is likely driven by broader geographic and climatic differences.
Regions with warmer climates, such as southern China, tend to have both higher vegetation and higher temperatures.
As a result, the increase in temperature across NDVI groups may reflect these underlying conditions rather than a direct effect of vegetation itself.

Overall, this distribution-based analysis provides a consistent picture with the earlier regression results and highlights that the relationship between NDVI and temperature is not simply a cooling effect, but is closely linked to regional and climatic context.

## 6.5 Change Analysis

To better capture dynamic relationships, we analyze year-to-year changes in NDVI and temperature.
First differences are calculated for both variables, and the relationship between changes in vegetation and changes in temperature is examined.
This analysis provides additional insight into whether increases or decreases in vegetation are associated with corresponding changes in temperature over time.

```{r}
# 1. Calculate year-to-year changes within each city
change_data <- data_clean %>%
  arrange(city, year) %>%
  group_by(city) %>%
  mutate(
    d_ndvi = ndvi - lag(ndvi),
    d_temperature = temperature - lag(temperature)
  ) %>%
  ungroup() %>%
  drop_na(d_ndvi, d_temperature)

glimpse(change_data)
summary(change_data)
```

```{r}
# 2. Scatter plot of year-to-year changes
ggplot(change_data, aes(x = d_ndvi, y = d_temperature)) +
  geom_point(alpha = 0.25, size = 1.6) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = "Year-to-Year Changes in NDVI and Temperature",
    x = "Change in NDVI",
    y = "Change in Temperature (°C)"
  )
```

The scatter plot shows the relationship between year-to-year changes in NDVI and temperature within the same city.
Compared to the earlier scatter plot, the points are now much more concentrated around zero, which makes sense because most year-to-year changes are relatively small.

Despite this, there is still a clear upward trend.
When NDVI increases in a given year, temperature also tends to increase in that same year.
The fitted line has a positive slope, and this pattern appears consistently across the data.

At the same time, the plot also shows some extreme observations, where both NDVI and temperature change sharply.
These outliers suggest that there may be occasional shocks or measurement noise, but they do not change the overall pattern.

```{r}
# 3. Correlation test for annual changes
change_cor_test <- cor.test(change_data$d_ndvi, change_data$d_temperature)
change_cor_test
```

The correlation coefficient is about 0.56, indicating a moderate positive relationship between changes in NDVI and changes in temperature.
The p-value is extremely small (p \< 0.001), which means this relationship is statistically significant.

Compared to the earlier level-based analysis, the correlation here is weaker but still clearly positive.
This suggests that even after focusing only on within-city changes, NDVI and temperature still tend to move in the same direction.

```{r}
# 4. Linear regression using first differences
change_lm <- lm(d_temperature ~ d_ndvi, data = change_data)
summary(change_lm)
```

The regression results confirm this pattern.
The coefficient on the change in NDVI is positive and highly significant, indicating that increases in NDVI are associated with increases in temperature within the same city over time.

In practical terms, the coefficient suggests that a one-unit increase in NDVI is associated with a large increase in temperature.
While the magnitude should be interpreted carefully, the direction of the relationship is very clear.

The R-squared is about 0.31, which means that changes in NDVI explain a meaningful portion of the variation in temperature changes, although a large share is still driven by other factors.

## 6.6 Spatial Visualization

To complement the statistical analysis, we incorporate a spatial perspective.
Maps are used to visualize the geographic distribution of NDVI and temperature across cities.
These spatial patterns help identify regional clusters and highlight geographic variation that may not be apparent in non-spatial plots.

```{r}
# 1. Load spatial data
china_city_sf <- st_read(
  here("data", "China city level shp.shp")
)

# Inspect spatial data
glimpse(china_city_sf)
st_crs(china_city_sf)
names(china_city_sf)
```

```{r}
# 2. Prepare one-year city-level data for mapping
# Here we use 2024 as the most recent year in the dataset
map_data_2024 <- data_clean %>%
  filter(year == 2024) %>%
  select(city_id, city, province, year, ndvi, temperature, region)
```

```{r}
# 3. Join city-level NDVI and temperature data to city boundary shapefile
china_city_sf <- china_city_sf %>%
  mutate(city_id = as.integer(city))

map_data_2024 <- data_clean %>%
  filter(year == 2024) %>%
  mutate(city_id = as.integer(city_id)) %>%
  select(city_id, city, province, year, ndvi, temperature, region)

china_map_2024 <- china_city_sf %>%
  left_join(map_data_2024, by = "city_id")
```

```{r}
# 4.Spatial Distribution of NDVI
ggplot(china_map_2024) +
  geom_sf(aes(fill = ndvi), color = NA) +
  labs(
    title = "Spatial Distribution of NDVI Across Chinese Cities, 2024",
    fill = "NDVI"
  ) +
  theme_void()
```

The NDVI map shows a clear spatial gradient across China.
Higher NDVI values are concentrated in southern and southeastern regions, especially in South China and parts of Southwest China.
These areas appear consistently darker on the map, indicating denser vegetation.

In contrast, northern and northwestern regions exhibit much lower NDVI levels.
Areas in Northwest China and parts of North China are visibly lighter, suggesting sparse vegetation coverage.
This pattern aligns well with known climatic and geographic differences, where southern China has more favorable conditions for vegetation growth, while arid and semi-arid regions in the northwest have limited vegetation.

```{r}
# 5. Spatial Distribution of Temperature
ggplot(china_map_2024) +
  geom_sf(aes(fill = temperature), color = NA) +
  labs(
    title = "Spatial Distribution of Temperature Across Chinese Cities, 2024",
    fill = "Temperature (°C)"
  ) +
  theme_void()
```

The temperature map reveals a spatial pattern that closely mirrors the NDVI distribution.
Warmer temperatures are concentrated in southern regions, particularly in South China, where average temperatures are the highest.

In contrast, northern and northeastern regions are much cooler, with noticeably lower temperature values.
The Northwest also shows relatively low temperatures compared to the south.

This spatial alignment between NDVI and temperature is striking.
Regions with higher NDVI tend to also have higher temperatures, while regions with lower NDVI tend to be cooler.
This provides a clear geographic explanation for the positive relationship observed in earlier analyses.

```{r}
# 6.Regional Average Map
region_summary_2024 <- map_data_2024 %>%
  group_by(region) %>%
  summarise(
    mean_ndvi = mean(ndvi, na.rm = TRUE),
    mean_temperature = mean(temperature, na.rm = TRUE),
    n_cities = n(),
    .groups = "drop"
  )

region_summary_2024
```

The regional summary statistics further support the spatial patterns observed in the maps.

South China has the highest average NDVI (0.67) and also the highest average temperature (23.48°C), making it both the greenest and warmest region.
Similarly, East China and Central China show relatively high NDVI levels and relatively high temperatures.

In contrast, Northwest China has the lowest average NDVI (0.30) and relatively low temperatures (7.43°C), while Northeast China also shows both low NDVI and low temperature.
North China falls in between but still follows the same general pattern.

These results confirm that NDVI and temperature tend to move together across regions.
The consistency between spatial maps and regional averages provides strong evidence that the observed relationship is driven by large-scale geographic differences rather than a simple vegetation effect.

# 7. Results and Interpretation

This section presents the main empirical findings derived from the exploratory data analysis (Section 5) and statistical modeling (Section 6). The results combine descriptive statistics, temporal trend analysis, distributional examination, and regression-based inference to provide a comprehensive understanding of NDVI dynamics and their relationship with temperature across Chinese cities. Overall, the analysis reveals three consistent patterns. First, both NDVI and temperature exhibit statistically significant upward trends over the period 2000–2024. Second, NDVI and temperature are positively associated, as supported by both correlation and regression results. Third, this relationship is not uniform across space, with substantial regional heterogeneity in both magnitude and direction. In addition, distributional and city-level analyses highlight considerable variability across observations, while year-to-year changes suggest that the NDVI–temperature relationship is more pronounced over long-term trends than short-term fluctuations. The following subsections present these findings in detail.

## 7.1 Temporal Trends in NDVI and Temperature

The temporal analysis indicates a clear upward trend in both NDVI and temperature across Chinese cities over the period 2000–2024. As illustrated in the annual average time series, mean NDVI increased steadily from approximately 0.44 in 2000 to above 0.51 by 2024, suggesting a gradual improvement in vegetation greenness over time. This pattern is further supported by the Mann–Kendall test, which identifies a statistically significant monotonic increase (z = 6.19, p \< 0.001, τ = 0.887).

A similar upward trajectory is observed for temperature. Mean temperature rose from around 13.0 °C in the early 2000s to above 14.2 °C in recent years. Although the series exhibits short-term fluctuations, including a noticeable dip around 2010–2012, the overall increasing trend is statistically significant (z = 3.53, p \< 0.001, τ = 0.507). Taken together, these results indicate that both vegetation activity and temperature have increased over the study period, with NDVI displaying a more consistent trend than temperature.

## 7.2 Distributional Characteristics of NDVI and Temperature

The distributional analysis provides additional insight into the variability of NDVI and temperature across all observations (n = 9,350). Summary statistics indicate that NDVI ranges from 0.060 to 0.829, with a mean of 0.478 and a median of 0.495. The interquartile range (0.388–0.610) suggests that most observations fall within moderate vegetation levels. The histogram and density plots show a multimodal distribution, with noticeable peaks around approximately 0.45–0.65, indicating heterogeneity in vegetation conditions across cities.

Temperature exhibits a wider spread, ranging from −5.604 °C to 27.053 °C, with a mean of 13.486 °C and a median of 15.140 °C. The interquartile range (8.704–17.728 °C) indicates substantial variation across climatic conditions. The distribution is also multimodal, with a dominant peak around 15–18 °C and smaller peaks at lower temperature ranges, reflecting the diversity of climatic zones across China.

Boxplot analysis further highlights differences in dispersion between the two variables. NDVI values are relatively tightly clustered, with most observations concentrated between approximately 0.38 and 0.61, and no strong evidence of extreme outliers. In contrast, temperature shows a wider spread and the presence of several lower-end outliers (below 0 °C), which are consistent with colder regions or seasonal variability captured in the dataset.

## 7.3 City-Level Variation

The dataset includes 369 cities, with a total of 9,350 observations. Most cities have complete time series coverage, with the number of years per city concentrated around 25 (median = 25, first quartile = 25, third quartile = 25), although the maximum reaches 50, indicating that a small number of cities may have duplicated or extended records.

City-level trajectory plots reveal substantial heterogeneity in both NDVI and temperature. While the overall trend line shows a gradual increase, individual city trajectories vary widely in both level and slope. For NDVI, values across cities span approximately from 0.1 to above 0.8, with most cities clustering between 0.3 and 0.7. For temperature, city-level values range from near 0 °C to above 25 °C, reflecting strong spatial variation across regions.

To better illustrate individual patterns, a subset of sampled cities is examined. These plots confirm that although most cities exhibit upward trends in both NDVI and temperature, the magnitude and variability differ significantly across locations. Some cities show relatively stable trajectories, while others display more pronounced fluctuations, highlighting localized dynamics.

## 7.4 Relationship Between NDVI and Temperature

The scatter plot analysis shows a clear positive association between NDVI and temperature across all observations. This relationship is statistically confirmed by the Pearson correlation test, which yields a correlation coefficient of 0.690 (95% confidence interval: 0.6797–0.7009, t = 92.284, df = 9348, p-value \< 2.2 × 10⁻¹⁶). This indicates a moderately strong and highly significant positive correlation between vegetation greenness and temperature.

The linear regression model further quantifies this relationship. The estimated coefficient for NDVI is 27.4477 (standard error = 0.2974, p-value \< 2 × 10⁻¹⁶), indicating a strong positive association. The intercept is 0.3605 (p = 0.0168). The model explains approximately 47.67% of the variance in temperature (R² = 0.4767), suggesting that NDVI is an important but not exclusive determinant of temperature variation. The residual standard error is 4.834, indicating moderate unexplained variability.

## 7.5 Regional Heterogeneity in the NDVI–Temperature Relationship

The relationship between NDVI and temperature varies substantially across regions. Faceted scatter plots show that regions such as East China, North China, and Southwest China exhibit strong positive relationships, while the Northeast region shows a negative slope. Central China and South China display relatively weak or flat relationships.

These differences are confirmed by the interaction regression model. The baseline NDVI coefficient is not statistically significant (coef = 0.6335, p = 0.5838), indicating that the overall relationship is not uniform when regional differences are accounted for. However, several interaction terms are highly significant. For example, NDVI has a strong positive effect in East China (coef = 11.4137, p \< 2 × 10⁻¹⁶), North China (coef = 14.0949, p \< 2 × 10⁻¹⁶), and especially Southwest China (coef = 39.6172, p \< 2 × 10⁻¹⁶). In contrast, the effect is negative in the Northeast (coef = −11.1180, p = 1.42 × 10⁻⁹). The model achieves a high explanatory power (R² = 0.8174), indicating that regional variation plays a critical role in shaping the NDVI–temperature relationship.

## 7.6 NDVI-Based Group Differences

Grouping observations by NDVI levels reveals a clear monotonic relationship between vegetation and temperature. The boxplots show that median temperature increases progressively from the low NDVI group to the high NDVI group. The lowest NDVI group corresponds to the lowest temperature distribution, while the highest NDVI group exhibits the highest median temperature.

Although there is variability within each group, the upward shift in both median and interquartile range across NDVI categories is consistent, reinforcing the positive association identified in the correlation and regression analyses.

## 7.7 Year-to-Year Changes

The analysis of first differences examines the relationship between annual changes in NDVI and temperature. The scatter plot shows a positive slope in the fitted linear trend, indicating that increases in NDVI are generally associated with increases in temperature at the year-to-year level. However, the data points are highly concentrated around zero, suggesting that most annual changes are relatively small.

At the same time, several extreme observations are present, with large positive or negative changes in either variable. Overall, the relationship in first differences appears weaker and more dispersed compared to the level-based analysis, indicating that the NDVI–temperature association is more clearly observed over long-term trends rather than short-term fluctuations.

# 8. Discussion

This section interprets the empirical findings in relation to the central research question: How does vegetation cover (NDVI) affect temperature across Chinese cities over time? Building on the results from Section 5 and Section 6, the discussion examines whether vegetation is associated with cooling or warming effects at the city-year scale, and how this relationship varies across temporal, spatial, and analytical dimensions. In particular, the analysis addresses the key sub-questions by evaluating long-term trends, cross-sectional associations, regional heterogeneity, differences across vegetation levels, and short-term dynamics. By situating the empirical results within existing literature, this section aims to clarify the mechanisms underlying the observed NDVI–temperature relationship and to assess the extent to which aggregated patterns reflect underlying environmental processes.

## 8.1 NDVI and Temperature Have Increased Concurrently Across Chinese Cities

The results show that both NDVI and temperature have increased significantly across Chinese cities from 2000 to 2024. This finding is consistent with two well-documented large-scale processes in China: climate warming and widespread vegetation greening. Previous studies have shown that China has experienced rapid warming over recent decades, particularly since the late 20th century (Piao et al., 2010). At the same time, satellite-based analyses indicate that China has contributed substantially to global vegetation greening(Zhu et al., 2016), driven by both climate change and large-scale ecological restoration programs (Chen et al., 2019).

The simultaneous increase in NDVI and temperature observed in this study therefore reflects broader coupled dynamics rather than independent trends. Rising temperatures may enhance vegetation growth by extending growing seasons and increasing photosynthetic activity in some regions (Piao et al., 2006), while policy-driven afforestation and land management programs have further amplified vegetation expansion (Liu et al., 2008). These processes together explain why NDVI and temperature increase in parallel over time.

## 8.2 NDVI Is Positively Associated with Temperature at the National Scale Rather Than Exhibiting a Cooling Effect

The empirical results do not support a simple negative relationship between vegetation and temperature at the national scale. Instead, the analysis consistently shows a positive association between NDVI and temperature across Chinese cities.

This finding may appear to contradict the commonly accepted cooling effect of vegetation, which operates through shading and evapotranspiration (Bowler et al., 2010). However, this apparent contradiction can be explained by the scale of analysis. At large spatial scales, temperature is often a key driver of vegetation productivity, meaning that warmer regions tend to have higher NDVI values (Nemani et al., 2003). As a result, the observed positive relationship primarily reflects how climate influences vegetation, rather than how vegetation modifies temperature.

In other words, the results suggest that temperature exerts a stronger influence on NDVI than NDVI exerts on temperature at the city-year scale. This distinction between climatic control and local cooling effects is widely recognized in the literature (Zhou et al., 2004).

## 8.3 The NDVI–Temperature Relationship Varies Substantially Across Regions

The results clearly show that the NDVI–temperature relationship varies substantially across regions. Some regions exhibit strong positive relationships, while others show weak or even negative associations.

This spatial heterogeneity is consistent with previous studies showing that vegetation–climate interactions depend heavily on regional environmental conditions (Zhang et al., 2001). In colder regions such as Northeast China, vegetation growth may be constrained by temperature, leading to different response patterns compared to warmer regions. In contrast, in humid and subtropical regions, vegetation may already be abundant, and additional temperature increases may not translate into the same vegetation response.

Urbanization may also contribute to these regional differences. Studies have shown that urban heat island effects can alter the relationship between vegetation and temperature, while vegetation distribution is shaped by land-use patterns and urban planning (Zhou et al., 2004). As a result, the observed NDVI–temperature relationship reflects a combination of climatic gradients, ecological processes, and human activities.

## 8.4 Higher NDVI Levels Are Associated with Higher Temperatures Due to Underlying Climatic Gradients

The NDVI group analysis indicates that temperature increases systematically with higher vegetation levels. While this pattern may appear inconsistent with the widely documented cooling effects of vegetation, it is more plausibly explained by underlying climatic gradients rather than direct causal effects of vegetation on temperature.

At large spatial scales, vegetation distribution is strongly constrained by climate, particularly temperature and water availability. Warmer and more humid regions tend to support higher vegetation productivity, while colder or arid regions exhibit lower NDVI values. As a result, NDVI and temperature often co-vary spatially due to shared environmental drivers. This interpretation is consistent with global-scale studies demonstrating that terrestrial vegetation activity is closely linked to climate variability, especially temperature and precipitation (Nemani et al., 2003).

Furthermore, studies on vegetation–climate interactions have shown that vegetation responses are not uniform but vary across environmental gradients and ecosystems (Seddon et al., 2016). Therefore, the positive association observed in this study is best understood as a reflection of large-scale climatic controls on vegetation, rather than evidence of a direct warming effect of vegetation.

## 8.5 The NDVI–Temperature Relationship Is Stronger Over Long-Term Trends Than Short-Term Changes

The first-difference analysis reveals a weaker and more dispersed relationship between changes in NDVI and changes in temperature, suggesting that the association is less pronounced at short temporal scales. This contrast indicates that the NDVI–temperature relationship is more robust in long-term trends than in year-to-year variability.

This finding aligns with previous research showing that vegetation–climate relationships are more clearly observed over longer time scales, while short-term dynamics are influenced by transient climatic variability and disturbances. For example, Piao et al. (2010) highlight that interannual variability in vegetation activity is strongly affected by short-term climate fluctuations, which may obscure longer-term relationships. Similarly, Pettorelli et al. (2005) note that NDVI is sensitive to seasonal and interannual variation, including phenological changes, which may not directly reflect structural shifts in vegetation cover.

As a result, the weaker relationship observed in first differences suggests that NDVI and temperature co-evolve over time primarily through long-term processes, while short-term interactions are more variable and less predictable.

## 8.6 Vegetation Cooling Effects Are Not Captured at the City-Year Scale of Analysis

Taken together, the results suggest that vegetation cover, as measured by NDVI, does not exhibit a uniform cooling effect on temperature across Chinese cities at the city-year scale. Instead, the observed positive association reflects the dominance of large-scale climatic gradients in the aggregated data.

This finding highlights the importance of spatial scale in interpreting vegetation–temperature interactions. Numerous studies have demonstrated that vegetation can reduce local temperatures through evapotranspiration and shading, but these effects are typically observed at fine spatial scales, such as neighborhood or land surface levels (Oke, 1982; Bowler et al., 2010). When data are aggregated to the level of entire cities and averaged annually, these localized cooling effects may be diluted or obscured.

In addition, the effectiveness of vegetation in mitigating heat is context-dependent and influenced by factors such as vegetation structure, spatial configuration, and the surrounding built environment (Gill et al., 2007; Ziter et al., 2019). NDVI, as a generalized indicator of vegetation greenness, does not capture these functional characteristics. Therefore, the results of this study should not be interpreted as evidence that vegetation lacks cooling effects, but rather that such effects are not detectable at the spatial and temporal resolution of the dataset used here.

# 9. Limitations

Textual Descriptions and Explanations

# 10. Conclusion

This study investigates the relationship between vegetation cover, as measured by NDVI, and temperature across Chinese cities over the period 2000–2024. By integrating exploratory data analysis with statistical modeling, the findings provide a comprehensive assessment of how vegetation and temperature co-evolve over time and across space.

The results indicate that both NDVI and temperature have increased significantly over the study period and are positively associated at the national scale. Rather than supporting a uniform cooling effect of vegetation, this positive relationship suggests that NDVI and temperature co-vary in response to shared climatic and geographic conditions. In this context, higher vegetation levels are more commonly observed in warmer regions, reflecting large-scale environmental gradients rather than direct thermal effects of vegetation.At the same time, the analysis reveals substantial regional heterogeneity in the NDVI–temperature relationship, with both the magnitude and direction of the association varying across regions. This spatial variation highlights the importance of contextual factors in shaping vegetation–climate interactions. Furthermore, the comparison between level-based and first-difference analyses suggests that the observed relationship is more robust over long-term trends than over short-term fluctuations.

Overall, these findings underscore that the relationship between vegetation and temperature is scale-dependent and context-specific. While vegetation may contribute to localized cooling at finer spatial scales, such effects are not captured at the aggregated city-year level used in this study. The results therefore emphasize the need to carefully consider spatial and temporal scale when interpreting vegetation–temperature relationships and assessing the role of vegetation in urban climate dynamics.

# 11. References

<https://www.mybib.com/j/KnowingFragileSwan>

# 12. Appendix

```{r eval=FALSE, include=FALSE}

```
