---
title: "Does Urban Vegetation Mitigate High Temperatures Across Cities?"
author: "Shiyi Zheng & Juanying Huangfu"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load packages
library(tidyverse)
library(lubridate)
library(sf)
library(here)
library(ggplot2)
```

# 1. Introduction

China’s rapid urbanization over the past decades has significantly intensified the Urban Heat Island effect (UHI), leading to rising urban temperatures and increasing concerns over heat-related health risks, energy demand, and long-term urban sustainability. Empirical studies have shown that urban areas in China have experienced faster warming trends compared to surrounding rural regions, largely driven by land-use change and the expansion of impervious surfaces (Zhou et al., 2014).Urban vegetation, commonly measured using the Normalized Difference Vegetation Index (NDVI), has been widely recognized as an effective nature-based solution to mitigate urban heat. Vegetation can reduce surface and air temperatures through mechanisms such as shading and evapotranspiration, thereby improving thermal comfort and reducing heat exposure(Bowler et al., 2010). And remote sensing studies using NDVI have consistently found negative associations between vegetation cover and land surface temperature across urban environments(Peng et al., 2011).

However, despite this well-established relationship, the extent to which vegetation mitigates temperature across Chinese cities over time remains insufficiently quantified. Most existing studies focus on single cities or short time periods, limiting the ability to capture long-term dynamics and cross-city variability. Moreover, regional heterogeneity—such as differences in climate zones, urban form, and baseline vegetation coverage—may lead to substantial variation in the strength and direction of the NDVI–temperature relationship(Zhao et al., 2014).

To address these gaps, this study conducts a multi-city analysis across Chinese prefecture-level cities from 2000 to 2024, integrating NDVI and temperature data to examine both temporal trends and cross-sectional relationships. Specifically, the analysis explores (1) long-term changes in vegetation and temperature, (2) the overall association between NDVI and temperature, (3) regional differences in this relationship, and (4) whether changes in vegetation are associated with corresponding changes in temperature over time. By combining time series analysis, regression approaches, and spatial visualization, this study aims to provide a systematic comprehensive understanding of the role of urban vegetation in mitigating heat across rapidly urbanizing regions.

**Research Question:**

- How does vegetation cover (NDVI) affect temperature across Chinese cities over time?

**Sub-questions:**

- How have NDVI and temperature changed over time?
- Is higher vegetation associated with lower temperatures?
- Does the relationship vary across regions?
- How does temperature differ across vegetation levels?
- Are changes in vegetation associated with changes in temperature?

---

# 2. Hypotheses

# 3. Data

## 3.1 Data Sources

This study uses two main datasets.

Vegetation data are measured using annual NDVI for Chinese prefecture-level cities from 2000 to 2024. These data are derived from MODIS satellite products and aggregated to the city level.

Temperature data consist of annual average temperature at the city level over the same period. These data are obtained from gridded climate datasets and spatially aggregated to match administrative boundaries.

## 3.2 Data Processing

The NDVI data are originally based on monthly raster observations and are aggregated into annual averages. Both NDVI and temperature datasets are spatially averaged within prefecture-level city boundaries using administrative shapefiles.

The two datasets are then merged by city and year to construct a consistent panel dataset.

## 3.3 Final Dataset

The final dataset is a city–year panel covering the period 2000 to 2024.

Key variables include:

- NDVI (annual average vegetation index)
- Temperature (annual average, °C)
- City identifier
- Year
- Province

---

# 4. Data Wrangling

In this step, the NDVI and temperature datasets are cleaned, standardized, and merged into a unified panel dataset.

The main steps include:

- Standardizing variable formats such as city names and year
- Converting variables into appropriate data types
- Merging NDVI and temperature data by city and year
- Removing missing or inconsistent observations
- Ensuring a consistent time period across cities

The resulting dataset forms the basis for all subsequent analysis.

```{r}
# Data import and cleaning steps

```

# 5. Exploratory Data Analysis
## 5.1 Temporal Trends
Textual Descriptions and Explanations
```{r}

```

## 5.2 Distribution of Key Variables
Textual Descriptions and Explanations
```{r}

```

## 5.3 City-Level Variation
Textual Descriptions and Explanations
```{r}

```


# 6. Methods and Analysis
## 6.1 Time Series Analysis
Textual Descriptions and Explanations
```{r}

```

## 6.2 Correlation and Linear Regression
Textual Descriptions and Explanations
```{r}

```

## 6.3 Regional Comparison
Textual Descriptions and Explanations
```{r}

```

## 6.4 Distribution Analysis
Textual Descriptions and Explanations
```{r}

```

## 6.5 Change Analysis
Textual Descriptions and Explanations
```{r}

```

## 6.6 Spatial Visualization
Textual Descriptions and Explanations
```{r}

```


# 7. Results and Interpretation
Textual Descriptions and Explanations

# 8. Discussion
Textual Descriptions and Explanations

# 9. Limitations
Textual Descriptions and Explanations

# 10. Conclusion
Textual Descriptions and Explanations

# 11. References
https://www.mybib.com/j/KnowingFragileSwan

# 12. Appendix
```{r eval=FALSE, include=FALSE}

```



