---
title: "Does Urban Vegetation Mitigate High Temperatures Across Cities?"
author: "Shiyi Zheng & Juanying Huangfu"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load packages
library(tidyverse)
library(lubridate)
library(sf)
library(here)
library(ggplot2)
```

# 1. Introduction

China’s rapid urbanization over the past decades has significantly intensified the Urban Heat Island effect (UHI), leading to rising urban temperatures and increasing concerns over heat-related health risks, energy demand, and long-term urban sustainability. Empirical studies have shown that urban areas in China have experienced faster warming trends compared to surrounding rural regions, largely driven by land-use change and the expansion of impervious surfaces (Zhou et al., 2014).Urban vegetation, commonly measured using the Normalized Difference Vegetation Index (NDVI), has been widely recognized as an effective nature-based solution to mitigate urban heat. Vegetation can reduce surface and air temperatures through mechanisms such as shading and evapotranspiration, thereby improving thermal comfort and reducing heat exposure(Bowler et al., 2010). And remote sensing studies using NDVI have consistently found negative associations between vegetation cover and land surface temperature across urban environments(Peng et al., 2011).

However, despite this well-established relationship, the extent to which vegetation mitigates temperature across Chinese cities over time remains insufficiently quantified. Most existing studies focus on single cities or short time periods, limiting the ability to capture long-term dynamics and cross-city variability. Moreover, regional heterogeneity—such as differences in climate zones, urban form, and baseline vegetation coverage—may lead to substantial variation in the strength and direction of the NDVI–temperature relationship(Zhao et al., 2014).

To address these gaps, this study conducts a multi-city analysis across Chinese prefecture-level cities from 2000 to 2024, integrating NDVI and temperature data to examine both temporal trends and cross-sectional relationships. Specifically, the analysis explores (1) long-term changes in vegetation and temperature, (2) the overall association between NDVI and temperature, (3) regional differences in this relationship, and (4) whether changes in vegetation are associated with corresponding changes in temperature over time. By combining time series analysis, regression approaches, and spatial visualization, this study aims to provide a systematic comprehensive understanding of the role of urban vegetation in mitigating heat across rapidly urbanizing regions.

**Research Question:**

- How does vegetation cover (NDVI) affect temperature across Chinese cities over time?

**Sub-questions:**

- How have NDVI and temperature changed over time?
- Is higher vegetation associated with lower temperatures?
- Does the relationship vary across regions?
- How does temperature differ across vegetation levels?
- Are changes in vegetation associated with changes in temperature?

---

# 2. Hypotheses

# 3. Data

## 3.1 Data Sources

This study uses two main datasets.

Vegetation data are measured using annual NDVI for Chinese prefecture-level cities from 2000 to 2024. These data are derived from MODIS satellite products and aggregated to the city level.

Temperature data consist of annual average temperature at the city level over the same period. These data are obtained from gridded climate datasets and spatially aggregated to match administrative boundaries.

## 3.2 Data Processing

The NDVI data are originally based on monthly raster observations and are aggregated into annual averages. Both NDVI and temperature datasets are spatially averaged within prefecture-level city boundaries using administrative shapefiles.

The two datasets are then merged by city and year to construct a consistent panel dataset.

## 3.3 Final Dataset

The final dataset is a city–year panel covering the period 2000 to 2024.

Key variables include:

- NDVI (annual average vegetation index)
- Temperature (annual average, °C)
- City identifier
- Year
- Province

---

# 4. Data Wrangling

In this step, the NDVI and temperature datasets are cleaned, standardized, and merged into a unified panel dataset.

The main steps include:

- Standardizing variable formats such as city names and year
- Converting variables into appropriate data types
- Merging NDVI and temperature data by city and year
- Removing missing or inconsistent observations
- Ensuring a consistent time period across cities

The resulting dataset forms the basis for all subsequent analysis.

```{r}
# Data import and cleaning steps

```

# 5. Exploratory Data Analysis
## 5.1 Temporal Trends
Textual Descriptions and Explanations
```{r}

```

## 5.2 Distribution of Key Variables
Textual Descriptions and Explanations
```{r}

```

## 5.3 City-Level Variation
Textual Descriptions and Explanations
```{r}

```


# 6. Methods and Analysis
## 6.1 Time Series Analysis
Textual Descriptions and Explanations
```{r}

```

## 6.2 Correlation and Linear Regression
Textual Descriptions and Explanations
```{r}

```

## 6.3 Regional Comparison
Textual Descriptions and Explanations
```{r}

```

## 6.4 Distribution Analysis
Textual Descriptions and Explanations
```{r}

```

## 6.5 Change Analysis
Textual Descriptions and Explanations
```{r}

```

## 6.6 Spatial Visualization
Textual Descriptions and Explanations
```{r}

```


# 7. Results and Interpretation
This section presents the main empirical findings derived from the exploratory data analysis (Section 5) and statistical modeling (Section 6). The results combine descriptive statistics, temporal trend analysis, distributional examination, and regression-based inference to provide a comprehensive understanding of NDVI dynamics and their relationship with temperature across Chinese cities. 

Overall, the analysis reveals three consistent patterns. First, both NDVI and temperature exhibit statistically significant upward trends over the period 2000–2024. Second, NDVI and temperature are positively associated, as supported by both correlation and regression results. Third, this relationship is not uniform across space, with substantial regional heterogeneity in both magnitude and direction. In addition, distributional and city-level analyses highlight considerable variability across observations, while year-to-year changes suggest that the NDVI–temperature relationship is more pronounced over long-term trends than short-term fluctuations. The following subsections present these findings in detail.

## 7.1 Temporal Trends in NDVI and Temperature
The temporal analysis indicates a clear upward trend in both NDVI and temperature across Chinese cities over the period 2000–2024. As illustrated in the annual average time series, mean NDVI increased steadily from approximately 0.44 in 2000 to above 0.51 by 2024, suggesting a gradual improvement in vegetation greenness over time. This pattern is further supported by the Mann–Kendall test, which identifies a statistically significant monotonic increase (z = 6.19, p < 0.001, τ = 0.887).

A similar upward trajectory is observed for temperature. Mean temperature rose from around 13.0 °C in the early 2000s to above 14.2 °C in recent years. Although the series exhibits short-term fluctuations, including a noticeable dip around 2010–2012, the overall increasing trend is statistically significant (z = 3.53, p < 0.001, τ = 0.507). Taken together, these results indicate that both vegetation activity and temperature have increased over the study period, with NDVI displaying a more consistent trend than temperature.

## 7.2 Distributional Characteristics of NDVI and Temperature
The distributional analysis provides additional insight into the variability of NDVI and temperature across all observations (n = 9,350). Summary statistics indicate that NDVI ranges from 0.060 to 0.829, with a mean of 0.478 and a median of 0.495. The interquartile range (0.388–0.610) suggests that most observations fall within moderate vegetation levels. The histogram and density plots show a multimodal distribution, with noticeable peaks around approximately 0.45–0.65, indicating heterogeneity in vegetation conditions across cities.

Temperature exhibits a wider spread, ranging from −5.604 °C to 27.053 °C, with a mean of 13.486 °C and a median of 15.140 °C. The interquartile range (8.704–17.728 °C) indicates substantial variation across climatic conditions. The distribution is also multimodal, with a dominant peak around 15–18 °C and smaller peaks at lower temperature ranges, reflecting the diversity of climatic zones across China.

Boxplot analysis further highlights differences in dispersion between the two variables. NDVI values are relatively tightly clustered, with most observations concentrated between approximately 0.38 and 0.61, and no strong evidence of extreme outliers. In contrast, temperature shows a wider spread and the presence of several lower-end outliers (below 0 °C), which are consistent with colder regions or seasonal variability captured in the dataset.

## 7.3 City-Level Variation
The dataset includes 369 cities, with a total of 9,350 observations. Most cities have complete time series coverage, with the number of years per city concentrated around 25 (median = 25, first quartile = 25, third quartile = 25), although the maximum reaches 50, indicating that a small number of cities may have duplicated or extended records.

City-level trajectory plots reveal substantial heterogeneity in both NDVI and temperature. While the overall trend line shows a gradual increase, individual city trajectories vary widely in both level and slope. For NDVI, values across cities span approximately from 0.1 to above 0.8, with most cities clustering between 0.3 and 0.7. For temperature, city-level values range from near 0 °C to above 25 °C, reflecting strong spatial variation across regions.

To better illustrate individual patterns, a subset of sampled cities is examined. These plots confirm that although most cities exhibit upward trends in both NDVI and temperature, the magnitude and variability differ significantly across locations. Some cities show relatively stable trajectories, while others display more pronounced fluctuations, highlighting localized dynamics.

## 7.4 Relationship Between NDVI and Temperature
The scatter plot analysis shows a clear positive association between NDVI and temperature across all observations. This relationship is statistically confirmed by the Pearson correlation test, which yields a correlation coefficient of 0.690 (95% confidence interval: 0.6797–0.7009, t = 92.284, df = 9348, p-value < 2.2 × 10⁻¹⁶). This indicates a moderately strong and highly significant positive correlation between vegetation greenness and temperature.

The linear regression model further quantifies this relationship. The estimated coefficient for NDVI is 27.4477 (standard error = 0.2974, p-value < 2 × 10⁻¹⁶), indicating a strong positive association. The intercept is 0.3605 (p = 0.0168). The model explains approximately 47.67% of the variance in temperature (R² = 0.4767), suggesting that NDVI is an important but not exclusive determinant of temperature variation. The residual standard error is 4.834, indicating moderate unexplained variability.

## 7.5 Regional Heterogeneity in the NDVI–Temperature Relationship
The relationship between NDVI and temperature varies substantially across regions. Faceted scatter plots show that regions such as East China, North China, and Southwest China exhibit strong positive relationships, while the Northeast region shows a negative slope. Central China and South China display relatively weak or flat relationships.

These differences are confirmed by the interaction regression model. The baseline NDVI coefficient is not statistically significant (coef = 0.6335, p = 0.5838), indicating that the overall relationship is not uniform when regional differences are accounted for. However, several interaction terms are highly significant. For example, NDVI has a strong positive effect in East China (coef = 11.4137, p < 2 × 10⁻¹⁶), North China (coef = 14.0949, p < 2 × 10⁻¹⁶), and especially Southwest China (coef = 39.6172, p < 2 × 10⁻¹⁶). In contrast, the effect is negative in the Northeast (coef = −11.1180, p = 1.42 × 10⁻⁹). The model achieves a high explanatory power (R² = 0.8174), indicating that regional variation plays a critical role in shaping the NDVI–temperature relationship.

## 7.6 NDVI-Based Group Differences
Grouping observations by NDVI levels reveals a clear monotonic relationship between vegetation and temperature. The boxplots show that median temperature increases progressively from the low NDVI group to the high NDVI group. The lowest NDVI group corresponds to the lowest temperature distribution, while the highest NDVI group exhibits the highest median temperature.

Although there is variability within each group, the upward shift in both median and interquartile range across NDVI categories is consistent, reinforcing the positive association identified in the correlation and regression analyses.

## 7.7 Year-to-Year Changes
The analysis of first differences examines the relationship between annual changes in NDVI and temperature. The scatter plot shows a positive slope in the fitted linear trend, indicating that increases in NDVI are generally associated with increases in temperature at the year-to-year level. However, the data points are highly concentrated around zero, suggesting that most annual changes are relatively small.

At the same time, several extreme observations are present, with large positive or negative changes in either variable. Overall, the relationship in first differences appears weaker and more dispersed compared to the level-based analysis, indicating that the NDVI–temperature association is more clearly observed over long-term trends rather than short-term fluctuations.




# 8. Discussion
This section interprets the empirical findings in relation to the central research question: How does vegetation cover (NDVI) affect temperature across Chinese cities over time? Building on the results from Section 5 and Section 6, the discussion examines whether vegetation is associated with cooling or warming effects at the city-year scale, and how this relationship varies across temporal, spatial, and analytical dimensions. In particular, the analysis addresses the key sub-questions by evaluating long-term trends, cross-sectional associations, regional heterogeneity, differences across vegetation levels, and short-term dynamics. By situating the empirical results within existing literature, this section aims to clarify the mechanisms underlying the observed NDVI–temperature relationship and to assess the extent to which aggregated patterns reflect underlying environmental processes.

## 8.1 NDVI and Temperature Have Increased Concurrently Across Chinese Cities
The results show that both NDVI and temperature have increased significantly across Chinese cities from 2000 to 2024. This finding is consistent with two well-documented large-scale processes in China: climate warming and widespread vegetation greening. Previous studies have shown that China has experienced rapid warming over recent decades, particularly since the late 20th century (Piao et al., 2010). At the same time, satellite-based analyses indicate that China has contributed substantially to global vegetation greening(Zhu et al., 2016), driven by both climate change and large-scale ecological restoration programs (Chen et al., 2019).

The simultaneous increase in NDVI and temperature observed in this study therefore reflects broader coupled dynamics rather than independent trends. Rising temperatures may enhance vegetation growth by extending growing seasons and increasing photosynthetic activity in some regions (Piao et al., 2006), while policy-driven afforestation and land management programs have further amplified vegetation expansion (Liu et al., 2008). These processes together explain why NDVI and temperature increase in parallel over time.

## 8.2 NDVI Is Positively Associated with Temperature at the National Scale Rather Than Exhibiting a Cooling Effect
The empirical results do not support a simple negative relationship between vegetation and temperature at the national scale. Instead, the analysis consistently shows a positive association between NDVI and temperature across Chinese cities.

This finding may appear to contradict the commonly accepted cooling effect of vegetation, which operates through shading and evapotranspiration (Bowler et al., 2010). However, this apparent contradiction can be explained by the scale of analysis. At large spatial scales, temperature is often a key driver of vegetation productivity, meaning that warmer regions tend to have higher NDVI values (Nemani et al., 2003). As a result, the observed positive relationship primarily reflects how climate influences vegetation, rather than how vegetation modifies temperature.

In other words, the results suggest that temperature exerts a stronger influence on NDVI than NDVI exerts on temperature at the city-year scale. This distinction between climatic control and local cooling effects is widely recognized in the literature (Zhou et al., 2004).

## 8.3 The NDVI–Temperature Relationship Varies Substantially Across Regions
The results clearly show that the NDVI–temperature relationship varies substantially across regions. Some regions exhibit strong positive relationships, while others show weak or even negative associations.

This spatial heterogeneity is consistent with previous studies showing that vegetation–climate interactions depend heavily on regional environmental conditions (Zhang et al., 2001). In colder regions such as Northeast China, vegetation growth may be constrained by temperature, leading to different response patterns compared to warmer regions. In contrast, in humid and subtropical regions, vegetation may already be abundant, and additional temperature increases may not translate into the same vegetation response.

Urbanization may also contribute to these regional differences. Studies have shown that urban heat island effects can alter the relationship between vegetation and temperature, while vegetation distribution is shaped by land-use patterns and urban planning (Zhou et al., 2004). As a result, the observed NDVI–temperature relationship reflects a combination of climatic gradients, ecological processes, and human activities.

## 8.4 Higher NDVI Levels Are Associated with Higher Temperatures Due to Underlying Climatic Gradients

The NDVI group analysis indicates that temperature increases systematically with higher vegetation levels. While this pattern may appear inconsistent with the widely documented cooling effects of vegetation, it is more plausibly explained by underlying climatic gradients rather than direct causal effects of vegetation on temperature.

At large spatial scales, vegetation distribution is strongly constrained by climate, particularly temperature and water availability. Warmer and more humid regions tend to support higher vegetation productivity, while colder or arid regions exhibit lower NDVI values. As a result, NDVI and temperature often co-vary spatially due to shared environmental drivers. This interpretation is consistent with global-scale studies demonstrating that terrestrial vegetation activity is closely linked to climate variability, especially temperature and precipitation (Nemani et al., 2003).

Furthermore, studies on vegetation–climate interactions have shown that vegetation responses are not uniform but vary across environmental gradients and ecosystems (Seddon et al., 2016). Therefore, the positive association observed in this study is best understood as a reflection of large-scale climatic controls on vegetation, rather than evidence of a direct warming effect of vegetation.

## 8.5 The NDVI–Temperature Relationship Is Stronger Over Long-Term Trends Than Short-Term Changes
The first-difference analysis reveals a weaker and more dispersed relationship between changes in NDVI and changes in temperature, suggesting that the association is less pronounced at short temporal scales. This contrast indicates that the NDVI–temperature relationship is more robust in long-term trends than in year-to-year variability.

This finding aligns with previous research showing that vegetation–climate relationships are more clearly observed over longer time scales, while short-term dynamics are influenced by transient climatic variability and disturbances. For example, Piao et al. (2010) highlight that interannual variability in vegetation activity is strongly affected by short-term climate fluctuations, which may obscure longer-term relationships. Similarly, Pettorelli et al. (2005) note that NDVI is sensitive to seasonal and interannual variation, including phenological changes, which may not directly reflect structural shifts in vegetation cover.

As a result, the weaker relationship observed in first differences suggests that NDVI and temperature co-evolve over time primarily through long-term processes, while short-term interactions are more variable and less predictable.

## 8.6 Vegetation Cooling Effects Are Not Captured at the City-Year Scale of Analysis
Taken together, the results suggest that vegetation cover, as measured by NDVI, does not exhibit a uniform cooling effect on temperature across Chinese cities at the city-year scale. Instead, the observed positive association reflects the dominance of large-scale climatic gradients in the aggregated data.

This finding highlights the importance of spatial scale in interpreting vegetation–temperature interactions. Numerous studies have demonstrated that vegetation can reduce local temperatures through evapotranspiration and shading, but these effects are typically observed at fine spatial scales, such as neighborhood or land surface levels (Oke, 1982; Bowler et al., 2010). When data are aggregated to the level of entire cities and averaged annually, these localized cooling effects may be diluted or obscured.

In addition, the effectiveness of vegetation in mitigating heat is context-dependent and influenced by factors such as vegetation structure, spatial configuration, and the surrounding built environment (Gill et al., 2007; Ziter et al., 2019). NDVI, as a generalized indicator of vegetation greenness, does not capture these functional characteristics. Therefore, the results of this study should not be interpreted as evidence that vegetation lacks cooling effects, but rather that such effects are not detectable at the spatial and temporal resolution of the dataset used here.


# 9. Limitations
Textual Descriptions and Explanations

# 10. Conclusion
Textual Descriptions and Explanations

# 11. References
https://www.mybib.com/j/KnowingFragileSwan

# 12. Appendix
```{r eval=FALSE, include=FALSE}

```



