---
title: "analysis_5_ExploratoryDataAnalysis"
author: "Shiyi Zheng & Juanying Huangfu"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load packages
library(tidyverse)
library(lubridate)
library(sf)
library(here)
library(ggplot2)

here()
```

# 5. Exploratory Data Analysis
## 5.1 Temporal Trends
As the first step in the exploratory data analysis, this study examines how NDVI and temperature evolve over time by aggregating city-level observations into annual averages. This transformation simplifies the panel dataset into a time series representation, allowing for a clear overview of long-term dynamics.

Line plots are used to visualize temporal changes in both variables, as they effectively capture year-to-year variation while preserving overall trends. To support interpretation, locally estimated scatterplot smoothing (LOESS) curves are applied to highlight underlying patterns by reducing short-term fluctuations.

The purpose of this step is not to establish relationships between variables, but to provide a descriptive overview of their temporal behavior. By identifying whether NDVI and temperature exhibit increasing, decreasing, or stable trends over the study period, this analysis lays the foundation for subsequent exploration of distributions (Section 5.2) and cross-city heterogeneity (Section 5.3).

```{r}
# 1. Read Data
library(readr)
processed_city_ndvi_temp_clean <- read_csv("data/processed/processed_city_ndvi_temp_clean.csv")

# 2. Check Data
View(processed_city_ndvi_temp_clean)
glimpse(processed_city_ndvi_temp_clean)
summary(processed_city_ndvi_temp_clean[, c("year", "ndvi", "temperature")])

#Check whether there are sufficient urban observations every year
processed_city_ndvi_temp_clean %>%
  group_by(year) %>%
  summarise(
    n_city = n_distinct(city),
    n_obs = n()
  )

```
The dataset contains city-level observations of NDVI and temperature from 2000 to 2024. Inspection of the data structure confirms that both NDVI and temperature are recorded as continuous numeric variables, alongside identifiers such as city and year.

Summary statistics show that NDVI values range from 0.060 to 0.829, with a mean of 0.478 and a median of 0.495. Temperature values range from -5.604°C to 27.053°C, with a mean of 13.486°C and a median of 15.140°C.

The yearly observation counts indicate a consistent number of cities (369) and total observations (374) across all years. This suggests that the dataset is balanced over time, with no missing years or substantial variation in the number of observations.
```{r}
# 3. Aggregate by year
ts_data <- processed_city_ndvi_temp_clean %>%
  group_by(year) %>%
  summarise(
    mean_ndvi = mean(ndvi, na.rm = TRUE),
    mean_temp = mean(temperature, na.rm = TRUE),
    .groups = "drop"
  )
```


```{r}
# 4. Temporal Trends of NDVI
ggplot(ts_data, aes(x = year, y = mean_ndvi)) +
  geom_line(color = "darkgreen", size = 1) +
  geom_smooth(method = "loess", color = "darkgrey", se = FALSE) +
  labs(
    title = "Temporal Trend of NDVI Across Cities in China (2000–2024)",
    x = "Year",
    y = "Mean NDVI",
  ) +
  theme_minimal()
```
The temporal trend of NDVI shows a clear and steady increase over the study period from 2000 to 2024. The mean NDVI rises from approximately 0.44 in 2000 to above 0.50 by 2024, indicating a gradual upward trend.

While minor year-to-year fluctuations are observed—particularly in the early 2000s and around 2008–2010—the overall trajectory remains consistently increasing. The LOESS smoothed curve further confirms this pattern, showing a continuous upward trend without major reversals.

Overall, the results suggest that average vegetation levels across cities have increased over time, with relatively stable growth throughout the study period.

```{r}
# 5. Temporal Trends of Temperature
ggplot(ts_data, aes(x = year, y = mean_temp)) +
  geom_line(color = "darkred", size = 1) +
  geom_smooth(method = "loess", color = "darkgrey", se = FALSE) +
  labs(
    title = "Temporal Trend of Temperature Across Cities in China (2000–2024)",
    x = "Year",
    y = "Mean Temperature (°C)",
  ) +
  theme_minimal()
```
The temporal trend of temperature shows an overall increasing pattern from 2000 to 2024. The mean temperature rises from approximately 13.0°C in 2000 to above 14.0°C by the end of the study period.

Compared to NDVI, temperature exhibits more pronounced year-to-year fluctuations, particularly between 2000 and 2012, where several short-term increases and decreases are observed. Despite this variability, the overall trend remains upward.

The LOESS smoothed curve highlights a relatively stable or slightly declining period in the early years, followed by a clearer and more consistent increase after around 2012. This indicates that the warming trend becomes more pronounced in the later period.

## 5.2 Distribution of Key Variables
Following the examination of temporal trends, the analysis next focuses on the distribution of key variables across all observations. Understanding the distributional properties of NDVI and temperature is essential for interpreting the overall data structure before conducting further analysis.

Histograms are used as a primary visualization tool to assess how values of NDVI and temperature are distributed across the dataset. These plots provide insights into the range, central tendency, and dispersion of each variable, while also revealing potential skewness or the presence of extreme values. In addition, summary statistics are used to complement the visual analysis by quantifying key characteristics such as mean, median, and variability.

This step serves as a diagnostic check on the dataset. It helps ensure that the variables behave as expected, identifies any potential outliers, and informs subsequent analyses, particularly regression and comparative analysis, where distributional assumptions and data balance may influence results.
```{r}
# 1. Reshape for combined exploratory plots
dist_data <- processed_city_ndvi_temp_clean %>%
  select(ndvi, temperature) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  )
```


```{r}
# 2. Histograms with density curves
ggplot(dist_data, aes(x = value, fill = variable)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 color = "white",
                 alpha = 0.6) +
  geom_density(color = "black", linewidth = 1) +
  facet_wrap(~ variable, scales = "free") +
  labs(
    title = "Distributions of NDVI and Temperature",
    x = "Value",
    y = "Density"
  ) +
  theme_minimal()
```
The distributions of NDVI and temperature exhibit distinct patterns across all observations. NDVI values are relatively concentrated within a narrower range, with most observations clustered between approximately 0.4 and 0.65. The density curve shows a slightly multimodal shape, indicating the presence of multiple concentration points within the mid-range values.

In contrast, temperature values display a wider spread, ranging from below 0°C to above 20°C. The distribution is more dispersed, with a dominant peak around the mid-teens, and a longer tail extending toward lower temperatures. The density curve also suggests multiple local peaks, reflecting variation across observations.

Overall, the comparison indicates that NDVI is more tightly distributed, while temperature exhibits greater variability and dispersion.

```{r}
# 3. Boxplots for dispersion and potential outliers
ggplot(dist_data, aes(x = variable, y = value, fill = variable)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red") +
  facet_wrap(~ variable, scales = "free_y")+
  labs(
    title = "Boxplots of NDVI and Temperature",
    x = "Variable",
    y = "Value"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
The boxplots provide a clearer view of the dispersion and range of NDVI and temperature values. NDVI values are relatively concentrated, with a narrow interquartile range centered around approximately 0.4 to 0.6. The median is close to the center of the box, indicating a fairly balanced distribution, and there are few visible outliers.

In contrast, temperature exhibits a much wider spread. The interquartile range spans a larger interval, and the overall range extends from below 0°C to above 25°C. Several lower-end outliers are observed, indicating the presence of relatively low temperature values within the dataset.

Overall, the boxplots confirm that temperature shows greater variability and dispersion than NDVI, while NDVI values remain more tightly clustered.

```{r}
# 4. Separate histogram for NDVI
ggplot(processed_city_ndvi_temp_clean, aes(x = ndvi)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 fill = "darkgreen",
                 color = "white",
                 alpha = 0.8) +
  geom_density(color = "black", linewidth = 1) +
  labs(
    title = "Distribution of NDVI Across All Observations",
    x = "NDVI",
    y = "Density"
  ) +
  theme_minimal()
```
The distribution of NDVI values is relatively concentrated within the mid-range, with most observations falling between approximately 0.4 and 0.65. The density curve indicates a slightly multimodal pattern, with two noticeable peaks within this interval.

Lower NDVI values (below 0.3) are less frequent, and only a small number of observations extend toward higher values above 0.7. Overall, the distribution suggests that most cities exhibit moderate levels of vegetation coverage, with limited extreme values.


```{r}
# 5. Separate histogram for Temperature
ggplot(processed_city_ndvi_temp_clean, aes(x = temperature)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 fill = "darkred",
                 color = "white",
                 alpha = 0.8) +
  geom_density(color = "black", linewidth = 1) +
  labs(
    title = "Distribution of Temperature Across All Observations",
    x = "Temperature (°C)",
    y = "Density"
  ) +
  theme_minimal()

```
The distribution of temperature values is more widely spread compared to NDVI, covering a broad range from below 0°C to above 20°C. Most observations are concentrated between approximately 8°C and 18°C, with a prominent peak around the mid-teens.

The density curve indicates a multimodal pattern, with additional smaller peaks at lower and higher temperature ranges. The left tail extends toward lower temperature values, suggesting the presence of relatively low-temperature observations, while a smaller right tail reaches above 20°C.

Overall, the distribution shows greater variability and dispersion than NDVI, with a wider range and more pronounced variation across observations.

## 5.3 City-Level Variation
Textual Descriptions and Explanations
```{r}

```
