---
title: "analysis_6_MethodsandAnalysis"
author: "Shiyi Zheng & Juanying Huangfu"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load packages
library(tidyverse)
library(lubridate)
library(sf)
library(here)
library(ggplot2)
library(trend)
```

# 6. Methods and Analysis

## 6.1 Time Series Analysis

This section examines long-term trends in NDVI and temperature over time. 
Following the logic of the course time series exercises, we first aggregate the panel data to yearly averages, then use line plots and smoothed trend lines to visualize temporal patterns. 
Because our dataset is annual rather than monthly or daily, the main focus here is long-term trend detection rather than seasonal decomposition.

```{r}
data_clean <- read.csv(
  here("data", "processed", "processed_city_ndvi_temp_clean.csv"),
  stringsAsFactors = FALSE
)

# Aggregate city-year panel data to annual mean values across all cities
annual_summary <- data_clean %>%
  group_by(year) %>%
  summarise(
    mean_ndvi = mean(ndvi, na.rm = TRUE),
    mean_temperature = mean(temperature, na.rm = TRUE),
    sd_ndvi = sd(ndvi, na.rm = TRUE),
    sd_temperature = sd(temperature, na.rm = TRUE),
    n_cities = n_distinct(city),
    .groups = "drop"
  )

annual_summary
```
The aggregated dataset summarizes the average vegetation and temperature conditions across Chinese prefecture-level cities for each year from 2000 to 2024.

```{r}
# 1. Smoothed trend plots
# NDVI Trend Over Time
ggplot(annual_summary, aes(x = year, y = mean_ndvi)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 1) +
  labs(
    title = "Average NDVI Across Chinese Cities, 2000–2024",
    x = "Year",
    y = "Mean NDVI"
  )

# Temperature Trend Over Time
ggplot(annual_summary, aes(x = year, y = mean_temperature)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 1) +
  labs(
    title = "Average Temperature Across Chinese Cities, 2000–2024",
    x = "Year",
    y = "Mean Temperature (°C)"
  )
```

The time series plots show a clear upward trend in both NDVI and temperature from 2000 to 2024, but the patterns are not identical.

NDVI increases quite steadily over time. There are small fluctuations in the early 2000s, but after around 2010 the increase becomes more consistent, and the overall trend is very smooth. This suggests that vegetation coverage in Chinese cities has been gradually improving over the past two decades.

Temperature also shows an overall increasing trend, but the pattern is more variable. There is a noticeable dip around 2010–2012, followed by a steady increase afterward. Compared to NDVI, temperature changes are less smooth and show more short-term fluctuations, which is expected given the influence of climate variability.

```{r}
# 2. Time series object representation
# Create annual time series objects
ndvi_ts <- ts(
  annual_summary$mean_ndvi,
  start = min(annual_summary$year),
  frequency = 1
)

temp_ts <- ts(
  annual_summary$mean_temperature,
  start = min(annual_summary$year),
  frequency = 1
)

# Plot time series objects
plot(
  ndvi_ts,
  main = "NDVI Time Series (Annual Average)",
  xlab = "Year",
  ylab = "Mean NDVI"
)

plot(
  temp_ts,
  main = "Temperature Time Series (Annual Average)",
  xlab = "Year",
  ylab = "Mean Temperature (°C)"
)
```

These time series plots show the same patterns in a more direct way, without smoothing. The upward trends in both NDVI and temperature are clearly visible, although year-to-year variation is more noticeable compared to the smoothed plots.

```{r}
# 3. Mann-Kendall Trend Tests
ndvi_mk <- mk.test(ndvi_ts)
temp_mk <- mk.test(temp_ts)

ndvi_mk
temp_mk
```
The Mann-Kendall test results support what we observe in the plots. NDVI shows a very strong and statistically significant increasing trend (z = 6.19, p < 0.001, τ = 0.89), indicating that the upward trend is highly consistent over time.

Temperature also shows a statistically significant increasing trend (z = 3.53, p < 0.001, τ = 0.51), although the strength of the trend is more moderate compared to NDVI.

One important takeaway is that both vegetation and temperature are increasing at the same time. This is directly relevant to our research question.

If vegetation has a cooling effect, we might expect higher NDVI to be associated with lower temperatures. However, the fact that both variables increase over time suggests that there may be broader factors, such as climate change, driving temperature upward.

This also highlights an important issue for the next step of the analysis: we need to separate the effect of vegetation from overall time trends, rather than relying only on simple correlations.

## 6.2 Correlation and Linear Regression: Relationship Between NDVI and Temperature

We next examine whether higher vegetation coverage is associated with lower temperature. 
Following the logic of the course GLM exercises, we first visualize the relationship using a scatter plot and a fitted linear trend line. 
We then use correlation analysis and a simple linear regression model to quantify the direction and strength of the association between NDVI and temperature.

```{r}
ndvi_temp <- data_clean %>%
  select(city, year, ndvi, temperature) %>%
  drop_na()

glimpse(ndvi_temp)
summary(ndvi_temp)
```

```{r}
# 1. Scatter plot with fitted linear trend
ggplot(ndvi_temp, aes(x = ndvi, y = temperature)) +
  geom_point(alpha = 0.3, size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(
    title = "Temperature and NDVI Across Chinese Cities",
    x = "NDVI",
    y = "Temperature (°C)"
  )
```

The scatter plot shows a clear positive relationship between NDVI and temperature. As NDVI increases, temperature also tends to increase, and the fitted regression line has a strong upward slope.

This pattern is quite striking. Instead of seeing a cooling effect of vegetation, the plot suggests that cities with higher NDVI tend to have higher temperatures on average. The points are fairly spread out, but the overall upward trend is very consistent across the full range of NDVI values.

```{r}
# 2.Pearson correlation test
cor_test_ndvi_temp <- cor.test(ndvi_temp$ndvi, ndvi_temp$temperature)
cor_test_ndvi_temp
```

The Pearson correlation coefficient is about 0.69, which indicates a relatively strong positive relationship between NDVI and temperature. The p-value is extremely small (p < 0.001), meaning this relationship is statistically significant and very unlikely to be due to random chance.

In practical terms, this means that NDVI and temperature tend to move together across observations. Cities and years with higher vegetation levels also tend to have higher temperatures.

```{r}
# 3.Simple linear regression
ndvi_temp_lm <- lm(temperature ~ ndvi, data = ndvi_temp)
summary(ndvi_temp_lm)
```

The regression results are consistent with the scatter plot and correlation test. The positive slope indicates that, on average, higher NDVI is associated with higher temperature.

This result is quite different from what we might expect if vegetation had a simple cooling effect. Instead of a negative relationship, we observe a strong positive association. This suggests that the relationship between vegetation and temperature is more complex than a simple “more trees, lower temperature” story.



## 6.3 Regional Comparison
Textual Descriptions and Explanations
```{r}

```

## 6.4 Distribution Analysis
Textual Descriptions and Explanations
```{r}

```

## 6.5 Change Analysis
Textual Descriptions and Explanations
```{r}

```

## 6.6 Spatial Visualization
Textual Descriptions and Explanations
```{r}

```
