---
title: "analysis_4_DataWrangling"
author: "Shiyi Zheng & Juanying Huangfu"
date: "2026-04-23"
output: pdf_document
toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load packages
library(tidyverse)
library(lubridate)
library(sf)
library(here)
library(ggplot2)

here()
```

# 4. Data Wrangling

In this step, the NDVI and temperature datasets are cleaned, standardized, and merged into a unified panel dataset.

The main steps include:

- Standardizing variable formats such as city names and year
- Converting variables into appropriate data types
- Merging NDVI and temperature data by city and year
- Removing missing or inconsistent observations
- Ensuring a consistent time period across cities

The resulting dataset forms the basis for all subsequent analysis.

```{r}
# Data import and cleaning steps

# check project root
here()

# create processed folder if it does not exist
dir.create(here("data", "processed"), showWarnings = FALSE, recursive = TRUE)

# 1. Read raw data

temp_raw <- read_csv(
  here("data", "raw", "clty level_temp_2000-2024.csv"),
  show_col_types = FALSE
)

ndvi_raw <- read_csv(
  here("data", "raw", "city level_ndvi_2000-2024.csv"),
  show_col_types = FALSE
)

# 2. Define year columns

year_cols <- as.character(2000:2024)

# 3. Reshape temperature data to long format

temp_long <- temp_raw %>%
  clean_names_basic() %>%
  pivot_longer(
    cols = any_of(year_cols),
    names_to = "year",
    values_to = "temperature"
  ) %>%
  mutate(
    year = as.integer(year),
    temperature = as.numeric(temperature)
  ) %>%
  arrange(city_id, year)

# 4. Reshape NDVI data to long format

ndvi_long <- ndvi_raw %>%
  clean_names_basic() %>%
  pivot_longer(
    cols = any_of(year_cols),
    names_to = "year",
    values_to = "ndvi"
  ) %>%
  mutate(
    year = as.integer(year),
    ndvi = as.numeric(ndvi)
  ) %>%
  arrange(city_id, year)


# 5. Check duplicate city-year observations

panel_check_temp <- temp_long %>%
  count(city_id, year) %>%
  filter(n > 1)

panel_check_ndvi <- ndvi_long %>%
  count(city_id, year) %>%
  filter(n > 1)

# 6. Merge temperature and NDVI

processed_city_panel <- temp_long %>%
  left_join(
    ndvi_long,
    by = c("city_id", "city", "pr_id", "province", "year")
  ) %>%
  mutate(
    temp_anomaly = temperature - mean(temperature, na.rm = TRUE),
    ndvi_z = as.numeric(scale(ndvi)),
    temperature_z = as.numeric(scale(temperature))
  ) %>%
  select(
    city_id, city, pr_id, province, year,
    ndvi, temperature, ndvi_z, temperature_z, temp_anomaly
  ) %>%
  arrange(city_id, year)


# 7. Summary before dropping missing values

missing_summary_before <- processed_city_panel %>%
  summarise(
    n_rows = n(),
    n_cities = n_distinct(city_id),
    year_min = min(year, na.rm = TRUE),
    year_max = max(year, na.rm = TRUE),
    missing_ndvi = sum(is.na(ndvi)),
    missing_temperature = sum(is.na(temperature))
  )

city_year_completeness_before <- processed_city_panel %>%
  group_by(city_id, city) %>%
  summarise(
    n_years = n(),
    missing_ndvi = sum(is.na(ndvi)),
    missing_temperature = sum(is.na(temperature)),
    .groups = "drop"
  )


# 8. Drop rows with missing values


processed_city_panel_clean <- processed_city_panel %>%
  drop_na(ndvi, temperature)


# 9. Summary after dropping missing values


missing_summary_after <- processed_city_panel_clean %>%
  summarise(
    n_rows = n(),
    n_cities = n_distinct(city_id),
    year_min = min(year, na.rm = TRUE),
    year_max = max(year, na.rm = TRUE),
    missing_ndvi = sum(is.na(ndvi)),
    missing_temperature = sum(is.na(temperature))
  )

city_year_completeness_after <- processed_city_panel_clean %>%
  group_by(city_id, city) %>%
  summarise(
    n_years = n(),
    .groups = "drop"
  )


# 10. Export processed files

write_csv(
  processed_city_panel_clean,
  here("data", "processed", "processed_city_ndvi_temp_clean.csv")
)

write_csv(
  missing_summary_before,
  here("data", "processed", "processed_city_ndvi_temp_summary_before_dropna.csv")
)

write_csv(
  missing_summary_after,
  here("data", "processed", "processed_city_ndvi_temp_summary_after_dropna.csv")
)

write_csv(
  city_year_completeness_before,
  here("data", "processed", "processed_city_ndvi_temp_completeness_before_dropna.csv")
)

write_csv(
  city_year_completeness_after,
  here("data", "processed", "processed_city_ndvi_temp_completeness_after_dropna.csv")
)

saveRDS(
  processed_city_panel_clean,
  here("data", "processed", "processed_city_ndvi_temp_clean.rds")
)

# 11. Print checks

print(missing_summary_before)
print(missing_summary_after)
print(panel_check_temp)
print(panel_check_ndvi)
```
